Governing through patronage: the bargain for education in decentralized Brazil

### 2021-03-30 paper/__latexindent_temp.tex
update images

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=blue,      
    urlcolor=blue,
    citecolor=blue
}
\captionsetup[figure]{font=small,labelfont=small}

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\renewcommand{\theproposition}{\arabic{proposition}}
\newcommand{\real}{\mathbb{R}_+^n}
\newcommand{\bfv}{\mathbf{v}}
\newcommand{\blue}{\textcolor{blue}}
% \renewcommand{\includegraphics}{\includegraphics}

@@

\begin{abstract}
    How can governments improve the quality of education? I argue that local politicians govern through patronage, undermining education at the expense of voters. Building on qualitative evidence in Brazil, I leverage large-scale administrative data to show that public education is captured by local political elites. Mayors buy off local legislators to enact policy agendas by offering them positions into the educational sector. The degree to which patronage occurs varies: when mayors have a stronger ally base in the city council, they face less pressure to bargain. Patronage induces turnover in educational staff, with negative downstream effects on student learning. Weak electoral backlash suggest that patronage is primarily a political elite game, with limited accountability to the electorate. These findings point to the dangers of elite capture of public services and its downstream consequences for social welfare.
\end{abstract}

@@

% what do I need to bring the audience onboard?
% emphasize the existence of discretion, turnover, political manipulation of public resources.
% in particular the existence of power-sharing arrangements between elected officials that go beyond electoral concerns

Across the world, vulnerable populations rely on governments for access to basic public services such as education. Yet in the developing world, the quality of public education remains low: functional illiteracy is widespread, and basic mathematical operations beyond the reach of many \citep{vegas_raising_2007}. A growing recognition of the importance of bureaucrats in explaining these outcomes \citep{pepinsky_bureaucracy_2017,finan_personnel_2015} has been accompanied by concern over the politician-bureaucrat nexus \citet{raffler_does_2016,gulzar_politicians_2017}. These actors, political and bureaucratic, have been extensively modeled under the dyadic framework of principal-agent models \citep{miller2005political}. Yet in most developing countries, these political and bureaucratic actors are multiple and the institutional environment under which they operate play a decisive role in how public services are managed and delivered.

In developing democracies, and within their political constituencies, politicians are seldom unitary. Elected officials, both executive and legislative, have at times overlapping or countervailing roles, with important implications as to how bureaucracies are staffed or the amount of discretion they receive \citep{huber_legislatures_2001,gailmard_formal_2012}. Where power-sharing arrangements exist, bureaucratic appointments reflect the relative power of particular constituencies \citep{francois_how_2015}. In the context of a separation of powers, the need for alignment between executive and legislative actors may require elected officials, in particular executive leaders who have the power to appoint staff into the bureaucracy, to form coalitions with legislators \citep{groseclose_1996_buying, banks_2000_buying, gottlieb_countervailing_2019}. Where programmatic commitments are weak, coalition-formation manifests itself as transactions in the form of patronage appointments to help secure votes by legislators \citep{power_optimism_2010, amorim_neto_presidential_2000}.

From an administrative perspective, individual bureaucrats are embedded in an administrative apparatus. And a key feature that distinguishes high performing bureaucratic institutions are relatively high levels of cohesion \citet{evans_embedded_1995,evans_bureaucracy_1999}. In other words, well-functioning bureaucracies are characterized by an ability to meritocratically recruit and retain talent. This is an organizational characteristic, which goes beyond any single bureaucrat and takes into account the bureaucracy as a whole. In contrast, bureaucracies exposed to high levels of turnover, in particular when the drivers of staff turnover are political, can display low levels of cohesion and morale. \citep{kohli_state-directed_2004, akhtari_political_2015}. Patronage transactions between elected officials, in particular when these result into a wave of new political appointees, disrupt established organizational procedures, throwing management into disarray and requiring an adjustment period to the new entries.

To shed light on this complexity, over the course of two years I conducted fieldwork research in municipalities across Brazil. In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}. However, the type of clientelism I observed was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government \citep{raile_executive_2011}. In Brazil, the exchange of public sector appointments for political favors -- a canonical example of patronage -- grants the mayor a carte blanche to enact her policy agenda.

In this paper, I argue that public education in municipalities across Brazil have been captured by political elites, with personnel appointments into public schools and management dominated by bargains between executive and legislative branches. Similar to coalition building in presidential contexts, mayors bargain for legislative support for their policy agenda through the allocation of public sector positions to city councilors and their constituencies \citep{laver_coalitions_1990, power_optimism_2010}. This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over voter welfare \citep{ferejohn_incumbent_1986}. By catering to the city council through patronage appointments, mayors shuffle the local educational bureaucracy, with negative downstream consequences for public school students.

The estimation proceeds in two parts. First, I show that political alignment between mayors and city councilors has a direct effect on patronage. Building on a canonical model of legislative vote-buying, I propose a theoretical model to analyze patronage under an institutionalized separation of powers, deriving my main empirical test: mayors who face less opposition in the city council engage in less patronage. To verify this hypothesis, I build a set of indicators to track educational staff turnover, leveraging administrative data of over 2 million school principals and teachers employed by municipalities. In line with qualitative accounts and theoretical expectations, mayors who have a stronger ally base in the legislature engage in less patronage. This results holds across a set of specifications and different measures of turnover.

I then show that this politically induced turnover has a negative impact on student learning. To estimate the effect of patronage on quality of education, I combine qualitative and quantitative evidence. Interviews conducted with educational bureaucrats and politicians suggested firm that turnover had a negative impact on teachers' ability to educate students, as time horizons were compressed and bonds between educators and learners were broken. To validate these accounts I combine administrative data on education from two separate student learning measures: \emph{Prova Brasil} and the \emph{SPAECE}. I construct multiple datasets to test these claims: the main specification contains over 1 million classrooms spread across the national territory. A set of estimations, combining multi-level modeling and fixed effects, provide strong evidence that teacher turnover has a negative effect on student learning.

This study contributes to an emerging literature on the politics of personnel and public services \citep{pepinsky_bureaucracy_2017, finan_personnel_2015}. I analyze how incentives shape political decision on how to manage local bureaucracies \citep{duflo_incentives_2012, gulzar_politicians_2017}, but I highlight the non-electoral policy goals of executive leaders. I also contribute to a growing research agenda on the consequences of political competition, demonstrating that increased political fragmentation between executive and legislative branches can lead patronage and decrease in the quality of public services. \citep{gottlieb_countervailing_2019, ferraz_corrupting_2012}. I bring to focus the end-to-end provision of public services, echoing a long-standing literature on state capacity \citep{kohli_state-directed_2004, evans_embedded_1995}.

The paper is structured as follows. Section \ref{sec:literature} provides an overview of the scholarly debate over public goods provision and personnel, as well as more specific treatments of the subject in Brazil. Section \ref{sec:context} presents the context and data. In section \ref{sec:theory}, I present the main argument, with a formal treatment of the subject using a canonical vote-buying model. Section \ref{sec:empirics} presents the research design and main results. Section 5 concludes.

\section{Related literature}
\label{sec:literature}

@@

% \subsection*{Presidential coalitionism and patronage}

% For every elected mayor in Brazil, a group of legislators are also elected into office. These political actors have competing claims over the local bureaucracy, with important implications for public service delivery. This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors \citep{grindle_jobs_2012, mccarty_appointments_2004}. Divergent political interests can lead to strategic interaction between executive and legislative actors. A rich literature in Brazil explores these relations, with important insights to how executive and legislators bargain over bureaucracy. \citet{raile_executive_2011, power_optimism_2010}.

% In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism. Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions. \citep{raile_executive_2011}. In a setting characterized by weak party cohesion and programmatic commitments \citet{ames_electoral_1995, lucas_ideological_2010}, bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda \citet{batista_o_2013, neto_presidential_2006, figueiredo_executivo_1999}.

% In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies. Due to weak programmatic commitments at the local level, public sector jobs are used to legislative support.\footnote{Interview with C, August and September 2019.} Mayors enjoy full discretion into how to appoint workers into the public sector, and use patronage to coopt legislative support from members of the coalition, a practice known locally as \textit{empreguismo}. As noted by a former mayor of the municipality of Sobral, "city councilors knocked on my door with a list of names for people they wanted me to hire."\footnote{Interview with C, August 2019.} These hires induce changes in personnel, with important consequences for bureaucracies and educational services.

@@

\section{Institutional context and data}
\label{sec:context}

@@

Brazil's local level of government is composed of over 5 thousand municipalities, each with their own city hall (\emph{prefeitura}) and city council (\emph{c\^{a}mara dos vereadores}). Mayors and city councilors (\textit{vereadores}) are democratically elected every four years, open to reelection for both offices.\footnote{The last municipal elections were in 2016. These take place every 4 years.} The executive is responsible for the management of public services such as education, with exclusive rights over public administration in general, and personnel appointment in particular \citep{akhtari_political_2015,brollo_victor_2017}. As noted by \citet{souza_reforma_2004}, decisions over whom to appoint into the educational sector are primarily in the hands of the mayor and her secretary of education. The city council, on the other hand, oversees legislation and approves the budget for the fiscal year, providing a local checks and balances on the executive.

At the federal level, this institutional arrangement gives rise to a phenomenon described as presidential coalitionism \citet{raile_executive_2011, power_optimism_2010}. Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions. \citep{raile_executive_2011}. In a setting characterized by weak party cohesion and programmatic commitments \citet{ames_electoral_1995, lucas_ideological_2010}, bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda \citet{batista_o_2013, neto_presidential_2006, figueiredo_executivo_1999}.

@@

Interviews with secretaries of education and mayors provide strong evidence that city councilors play an important role in staffing decisions, despite having no \emph{formal} rights to do so. While mayors hold jurisdiction over personnel decisions, there are extensive consultations between mayors and city councilors to decide who becomes the principal of a school, or which teacher remains in a school or leaves. These executive-legislative consultations ultimately determine the allocation (\textit{alotação}) of educational staff, serving as the primary tool through which mayors reward or coopt legislators into supporting them in the city council. According to accounts by school managers in the municipality of I:

@@

    \emph{Educational councilor}: I was also invited to work here, by the city councilor and the department of education.\footnote{Interview with the administrative board of school A., municipality J. August 2019.}
\end{quote}

In Brazil, the responsibility to provide public, primary education is primarily delegated to local governments \citet{paschoal_historia_2009}. In this critical learning period, students acquire skills such as reading/writing, as well as foundational concepts in mathematics. The municipal educational system has increased in size over the past few decades, and currently over 60 percent of lower school students attend a public school. Figure \ref{fig:enroll} plots the total number of students in primary education per government level, including private schools. As of 2016, over 25 million students were enrolled in over 115 thousand municipal schools.

\begin{figure}[h]
    \centering
    \includegraphics{plots/descriptive_enrollment.pdf}
    \caption{\textbf{Student enrollment by administration.}}
    \label{fig:enroll}
\end{figure}

Teachers and school principals are responsible for providing all educational services at the local level, overseen by a local department of education (\emph{secretaria de educa\c{c}\~{a}o}). As of 2016, in the municipal education system there were 3.7 million teachers distributed across the national territory, as well as 148 thousand school principals. A majority of teachers are hired with a permanent contract (\emph{estatutário}), with over 73 percent of teachers enrolled with civil service protection. This contrasts with the hiring of school principals, who are politically appointed and therefore have no similar protections set in place.

\begin{table}[t]
    \input{results/descriptive_teacher}
    \caption{\textbf{Summary statistics for municipal teachers.} All statistics take as reference year 2017.}
    \label{tbl:descriptive_teacher}
\end{table}

Table \ref{tbl:descriptive_teacher} provides some additional descriptive statistics. The teacher labor force is highly educated, with over 70 percent of teachers holding a higher education degree. The majority of teachers are women, and schools have a limited amount of teachers, with an average of 4 per school.

\begin{figure}[h]
    \centering
    \includegraphics{plots/plot_principal_descriptive.pdf}
\end{figure}

Positions in the educational sector are politically valuable. As a department, it comprises over 25 percent of local public sector jobs. Due to their relatively high compensation and social status compared to other positions in the local bureaucracy, educational staff positions are particularly valuable for constituencies seeking employment. Additionally, qualitative evidence collected during fieldwork suggests that teachers play an important role in local party networks, leading electoral campaigns and brokering votes, similar to dynamics found by \citet{oliveros_making_2016} in Argentina's campaigning teachers.

\begin{figure}[h]
    \centering
    \includegraphics{plots/descriptive_staff_breakdown}
\end{figure}

@@

Brazil collects fine-grained data on education and makes it publicly available for research. For this study, data on education are collected from two main sources: the SAEB and School Census.\footnote{These can be accessed \hyperlink{http://portal.inep.gov.br/web/guest/dados}{here}.} The SAEB (National System for Educational Assessment) is a biannual standardized exam administered by the INEP (National Institute for Educational Studies and Research) to all public schools and a sample of private schools. In 2017, over 5 million students, in 5th and 9th grade, took the exam, testing their proficiency in both mathematics and Portuguese. In this study, test scores are the primary metric for assessing the quality of education received by students.

\begin{figure}[h]
    \centering
    \includegraphics[]{plots/saeb_map.pdf}
    \caption{Uneven quality of municipal education. Polygons represent municipalities, warmer colors indicate higher municipal average in standardized test scores. Averages are weighted by student participation rate.}
    \label{fig:saeb_map}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics{plots/descriptive_staff_turnover.pdf}
    \caption{Proportion of public sector jobs by department. Note that I only keep the top 5 categorie}
    \label{fig:staff_turnover}
\end{figure}

@@

\blue{Insert more information and descriptive paragraph on turnover index here}

@@

\begin{figure}
    \centering
    \includegraphics{plots/hist_coalition.pdf}
    \caption{Distribution of share of legislative seats allied with the mayor during the electoral campaign for the elecoral period of 2002-2016.}
    \label{fig:hist_coalition}
\end{figure}

Additional data has been collected to supplement the estimation exercise. Sociodemographic data comes from the National Institute for Statistics and Geography (IBGE), budgetary data from (FINBRA) and student test scores from Ceará (SPAEEC) are graciously provided by the Department of Education of Ceará.\footnote{Respectively, data can be found here: (1)\hyperlink{https://ces.ibge.gov.br/base-de-dados/links-base-de-dados.html}{IBGE}, (2)\hyperlink{https://siconfi.tesouro.gov.br/siconfi/pages/public/consulta_finbra/finbra_list.jsf}{Finbra}, (3)\hyperlink{https://www.seduc.ce.gov.br/spaece/}{Spaece}.}

@@

\section{Empirical results}
\label{sec:empirics}

@@

This study presents two sets of estimations. The first one identifies the effect of bureaucratic turnover on student learning, verifying the hypothesis that teacher and school principal turnover has a negative effect on the quality of municipal education. To do so, I leverage the test scores made available by \emph{SAEB} and \emph{SPAECE} to estimate the effect of turnover on student test scores. To avoid interference, individual test scores are aggregated at the classroom level, for each 5th and 9th grade of each school contained in the sample. The outcome of interest is thus the average test scores for all students in the same classroom (SAEB) or same grade (SPAECE), for a given school.

@@

% Despite the use of different educational quality and staff turnover metrics, addition of a relatively extensive set of controls, the estimation strategy proposed here remains observational. In the absence of exogenous shock in either turnover or coalition shares, a credible identification strategy for each set of estimations remains a challenge. To partially address these concerns, as an additional robustness check I replicate these estimations with weights estimated through a covariate balancing propensity score outlined in \citet{imai_covariate_2014} which is generalized for continuous treatments. A covariate balance test for either estimation is presented in the appendix.

@@

\begin{figure}[h]
    \centering
    \includegraphics{plots/plot_coalition_coef_mun.pdf}
    \caption{{\bf Political alignment and staff turnover.} A visual representation of results presented in table \ref{tbl:turnover}. All models include year and state fixed effects.}
    \label{fig:coalition_turnover}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics{plots/plot_interactions_logit}
    \caption{\textbf{Differential reductions in patronage in teachers and school principals}. Predicted values for bureaucratic turnover plotted against the proportion of educational staff either hired or fired for any given year against the share of seats controlled by the executive.}
    \label{fig:turnover}
\end{figure}

Finally, I show that the executive-legislative bargain extends to second-term mayors. I subset the data used for the above analysis to only mayors who are reelected, and re-run the previous analyses with the same specification. I find that the results are similar in magnitude and precision to those presented above. This finding suggests that patronage is not motivated solely by reelection concerns. Rather, patronage achieves an important policy goal for incumbent mayors who seek to implement their preferred policy, regardless of whether they are just initiating their mandate or concluding it.

\begin{figure}[h]
    \centering
    \includegraphics{plots/plot_coalition_coef_mun_second.pdf}
    \caption{\textbf{Patronage continues into the second term.} There is strong evidence that executive-legislative bargains continue into the mayor's second term. All models include a full set of controls, as well as year and state fixed effects.}
\end{figure}

Moving on to characterizing the downstream effects of patronage, I find that teacher and school principal turnover have significant, negative effects on average test scores. These results are robust to alternative specifications of turnover, as well as the use of different quality metrics (SAEB, SPAECE) in our estimations. Table \ref{tbl:student_learning} presents the results of the estimation on SAEB and SPAECE test scores. We present two specifications for turnover. Turnover index measures the amount of turnover in teachers at the school level. Work experience for teachers and school principals serve as an alternative measure of turnover.

\begin{landscape}
    \begin{table}[t]
      \centering
      \footnotesize
      \input{results/student_learning.tex}
      \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning. Models 1 and 2 present results for teacher turnover index constructed at the school level. Models 3 and 4 estimate the effect of new teachers and school principals entering the school (less than two years). All models include year and state fixed effects.}
      \label{tbl:student_learning}
    \end{table}
\end{landscape}

I also present the coefficients for the regression exercise above in figure \ref{fig:hlm_mods}. Note that the results are consistent across the board, providing strong evidence that staff turnover has negative downstream consequences for student learning.

\begin{figure}[h]
  \centering
  \includegraphics{plots/model_turnover_learning.pdf}
  \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning, across different sets of exams (\emph{SAEB} and \emph{SPAECE}). All models include a full set of controls, as well as year and state fixed effects.}
  \label{fig:hlm_mods}
\end{figure}

Finally, I present some evidence of differential returns to patronage, in particular for city councilors. For each position (mayor, city councilor) I estimate the probability of being reelected conditional on the amount of patronage occurred in the first term manda While mayors themselves do not directly benefit from greater patronage, city councilors who are not members of the electoral base of the mayor seem to be negatively affected by patronage appointments into education. The negative consequences of patronage on student learning, however, do not bite, as nayors who deliver a worse quality of education are no less likely to be reelected in the next term.\footnote{See appendix \ref{app:accountability}.}

\begin{figure}[h]
    \centering
    \includegraphics{plots/reelection_marginal_effect_plots.pdf}
    \caption{\textbf{Differential returns to patronage}. Mayors who engage in patronage do not directly benefit from it. Instead, city councilors in the mayoral ally base seem to be benefiting from it at the expense of non-allied city councilors.}
\end{figure}

As a final note, this set of estimations are based on observational data and therefore suffer from well-known concerns of endogeneity. It is in these circumstances that validating causal mechanisms through fieldwork and careful data analysis can increase the validity of a causal claim. In-depth interviews with local actors involved in managing education, as well as the use of different measurements for both educational staff turnover and student learning, provides strong evidence that staff turnover has a negative impact on student learning. Bureaucratic turnover stems from political considerations.

@@

\section{Conclusion}
\label{sec:conclusion}

@@

\pagebreak

\pagebreak

\section*{Appendix}

\subsection{Covariate balance}

\begin{figure}[h]
    \centering
    \includegraphics{plots/covariate_balance.pdf}
\end{figure}

\pagebreak

\subsection{Accountability of mayors}
\label{app:accountability}

\begin{figure}[h]
    \centering
    \includegraphics{plots/accountability_fit.pdf}
\end{figure}

>>>>

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=blue,      
    urlcolor=blue,
    citecolor=blue
}

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\renewcommand{\theproposition}{\arabic{proposition}}
\newcommand{\real}{\mathbb{R}_+^n}
\newcommand{\bfv}{\mathbf{v}}
\newcommand{\blue}{\textcolor{blue}}
% \renewcommand{\includegraphics}{\includegraphics[width=0.6\linewidth]}

@@

\begin{abstract}
    Why do politicians engage in patronage and what are its consequences for public service delivery? I find that staff turnover in educational staff results from mayors attempt to exchange public sector jobs for support from local legislators. Leveraging fieldwork evidence and administrative data, I demonstrate that mayors with a weaker hold on the local legislature resort to patronage to co-opt legislators. Educational staff turnover has negative downstream effects on student learning, but limited electoral backlash. Weak electoral accountability channels suggests that patronage is primarily a political elite game, with little involvement by the electorate. These findings point to the dangers of elite capture of public services and its downstream consequences for social welfare.
\end{abstract}

@@

% Across the world, vulnerable populations rely on governments for access to basic public services such as education. In many countries, particularly in the developing world, the quality of public education remains woefully low: functional illiteracy remains widespread, mathematical operations, elusive. Clearly, politicians play an important role in delivering these services \citep{gulzar_politicians_2017,besley_political_2003}, but what are the incentives they face and what are their implications for bureaucratic institutions responsible for delivering them. \citep{cochrane_policies_1983, falleti_decentralization_2010}. How can local governments improve the quality the of public services they deliver? In particular, how do political actors reshape bureaucratic institutions, and what are their downstream consequences for student learning?

The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, and children do not master basic arithmetic operations. How can these educational outcomes be improved? Previous studies have found that individual incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}. Often relying on quasi-experimental research designs, these studies provide compelling evidence that political and bureaucratic actors respond to policy interventions.\citep{finan_personnel_2015} But it remains unclear how these incentives connect to the broader institutional environment: who delivers public services and what political institutions shape their behavior?

In this paper, I build on extensive fieldwork research across Brazil to argue that the executive-legislative bargain shapes public education, leading to capture by local political elites. Similar to coalition building in presidential contexts, mayors bargain for legislative support for their policy agenda through the allocation of public sector positions to city councilors and their constituencies \citep{laver_coalitions_1990, power_optimism_2010}. This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over voter welfare \citep{ferejohn_incumbent_1986}. By catering to the city council through patronage appointments, mayors ultimately cause instability in the local educational bureaucracy, with negative downstream consequences for public school students.

To analyze how politics affect staff turnover, I analyze the relationship between municipal education staff turnover and the share of seats controlled by the executive in the local city council. In line with qualitative accounts and theoretical expectations, mayors who face stronger legislative opposition resort to greater patronage, in particular during the first year of administration. Mayors therefore resort to patronage in order to shore up legislative support, with negative consequences for student learning. I leverage data on over 2 million school teachers and school principals to estimate the effect of share of executive seats on turnover. Different model specifications confirm that turnover increases (decreases) as the share of seats held by the mayor coalition decrease (increase).

To estimate the effect of patronage on quality of education, I combine qualitative and quantitative evidence. Interviews conducted with educational bureaucrats and politicians confirm that turnover has a negative impact on teachers' ability to educate students. To validate these accounts I combine administrative data on education from the \emph{Prova Brasil}, a nation-wide standardized exam designed to measure student learning. I construct a multiple datasets to test these claims: the main specification contains over 1 million classrooms spread across the national territory. Using the national school census (\emph{Censo Escolar}), I construct a school-specific turnover index that tracks year-by-year volatility in teacher staff. A set of estimations, combining multi-level modeling and fixed effects, provide strong evidence that teacher turnover has a negative effect on student learning.

This study contributes to an emerging literature on the politics of personnel and public services \citep{pepinsky_bureaucracy_2017, finan_personnel_2015, gulzar_politicians_2017}. Focusing on political actors refines our understanding of local politics and how intra-elite bargain reshape bureaucracies, leading to welfare loss for the broader population, as subnational actors divert resources from public services \citep{ferraz_corrupting_2012}. It also shifts focus from provision to the implementation of public services, echoing a long-standing literature on state capacity and the necessity for bureaucratic autonomy \citep{kohli_state-directed_2004, evans_embedded_1995}.

The paper is structured as follows. Section 1 provides an overview of the scholarly debate over public goods provision and personnel, as well as more specific treatments of the subject in Brazil. In section 2 I present the main theoretical argument, with a formal treatment of the subject. Section 3 presents the institutional context and data, followed in section 4 by the research design and main results. Section 5 concludes.

\section{Related literature}

@@

\subsection*{Presidential coalitionism and patronage}

For every elected mayor in Brazil, a group of legislators are also elected into office. These political actors have competing claims over the local bureaucracy, with important implications for public service delivery. This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors \citep{grindle_jobs_2012, mccarty_appointments_2004}. Divergent political interests can lead to strategic interaction between executive and legislative actors. A rich literature in Brazil explores these relations, with important insights to how executive and legislators bargain over bureaucracy. \citet{raile_executive_2011, power_optimism_2010}.

In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism. Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions. \citep{raile_executive_2011}. In a setting characterized by weak party cohesion and programmatic commitments \citet{ames_electoral_1995, lucas_ideological_2010}, bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda \citet{batista_o_2013, neto_presidential_2006, figueiredo_executivo_1999}.

@@

\section{Institutional context and data}

In Brazil, responsibility for primary education is delegated to local governments \citet{paschoal_historia_2009}. The municipal educational system has increased in relevance over the past few decades, as municipalities took on a larger proportion of students from state-level governments. Figure \ref{fig:enroll} plots the total number of students in primary education per government level, including private schools. As of 2016, over 25 million students were enrolled in over 115 thousand municipal schools.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{plots/descriptive_enrollment.pdf}
    \caption{enrollment}
    \label{fig:enroll}
\end{figure}

Positions in the educational sector are politically valuable. As a department, it comprises over 25 percent of local public sector jobs. Due to their relatively high compensation and social status compared to other positions in the local bureaucracy, educational staff positions are particularly valuable for constituencies seeking employment. Additionally, anecdotal evidence collected during fieldwork suggests that teachers play an important role in local party networks, leading electoral campaigns and brokering votes, similar to dynamics found by \citet{oliveros_making_2016} in Argentina's campaigning teachers.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{plots/descriptive_staff_breakdown}
\end{figure}

@@

Brazil's local level government is composed by over 5 thousand municipalities, each with their own mayor hall (\emph{prefeitura}) and city council (\emph{c\^{a}mara dos vereadores}). Mayors and city councillors (\textit{vereadores}) are democratically elected every four years, with the possibility of reelection for both.\footnote{The last municipal elections were in 2016. These take place every 4 years.} The executive is responsible for the management of public services such as education, with exclusive rights over personnel apppointment. The city council, on the other hand, oversees legislation and approves the budget for the fiscal year. As noted by \citet{souza_reforma_2004}, decisions over whom to appoint into the educational sector are primarily in the hands of the mayor and her secretary of education.

@@

Interviews with secretaries of education and mayors provide evidence that city councilors play an important role in staffing decisions. While mayors hold jurisdiction over personnel decisions, there are extensive consultations between mayors and city councilors to decide who becomes the principal of a school, or which teacher remains in a school or leaves. These executive-legislative consultations ultimately determine the allocation (\textit{alotação}) of educational staff, serving as the primary tool through which mayors reward or coopt legislators into supporting them in the city council. In the words of a set of school administrators in the municipality of I:

@@

    \emph{Educational counselor}: I was also invited to work here, by the city councilor and the department of education.\footnote{Interview with the administrative board of school A., municipality J. August 2019.}
\end{quote}

These accounts, along with previous case studies of municipalities in Brazil, suggest that city councilors play an important role in nominating staff. To verify these claims in a broader set of municipalities, as well as outlining the research design, I apply a theory of legislative vote-buying that illustrates the  employ extensive administrative and electoral data collected by the Brazilian federal government. In the next section I describe the data, where it is publicly available, as well as the preparation necessary for the set of estimations presented in the study.

@@

Brazil collects fine-grained data on education and makes it publicly available for research. For this study, data on education are collected from two main sources: the SAEB and School Census.\footnote{These can be accessed \hyperlink{http://portal.inep.gov.br/web/guest/dados}{here}.} The SAEB (National System for Educational Assessment) is a biannual standardized exam administered by the INEP (National Institute fosr Educational Studies and Research) to all public schools and a sample of private schools. In 2017, over 5 million students, in 5th and 9th grade, took the exam, testing their proficiency in both mathematics and Portuguese. In this study, test scores are the primary metric for assessing the quality of education received by students.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{plots/saeb_map.pdf}
    \caption{Uneven quality of municipal education. Polygons represent municipalities, warmer colors indicate higher municipal average in standardized test scores. Averages are weighted by student participation rate.}
    \label{fig:saeb_map}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.\linewidth]{plots/descriptive_staff_turnover.pdf}
    \label{fig:staff_turnover}
    \caption{Proportion of public sector jobs by department. Note that I only keep the top 5 categorie}
\end{figure}

@@

@@

\begin{figure}
    \centering
    \includegraphics[width=0.6\linewidth]{plots/hist_coalition.pdf}
    \caption{Distribution of share of legislative seats allied with the mayor during the electoral campaign for the elecoral period of 2002-2016.}
    \label{fig:hist_coalition}
\end{figure}

Additional data has been collected to supplement the estimation exercise. Sociodemographic data comes from the National Institute for Statistics and Geography (IBGE), budgetary data from (FINBRA) and student test scores from Ceará (SPAEEC) are graciously provided by the Department of Education of Ceará.\footnote{Respectively, data can be found here: (1)\url{https://ces.ibge.gov.br/base-de-dados/links-base-de-dados.html}, (2)\url{https://siconfi.tesouro.gov.br/siconfi/pages/public/consulta_finbra/finbra_list.jsf}, (3)\url{https://www.seduc.ce.gov.br/spaece/}.}

@@

\section{Empirical results}

@@

This study presents two sets of estimations. The first one identifies the effect of bureaucratic turnover on student learning, verifying the hypothesis that teacher and school principal turnover has a negative effect on the quality of municipal education. To do so, I leverage the test scores made available by SAEB and SPAECE to estimate the effect of turnover on student test scores. To avoid interference, individual test scores are aggregated at the classroom level, for each 5th and 9th grade of each school contained in the sample. The outcome of interest is thus the average test scores for all students in the same classroom (SAEB) or same grade (SPAECE), for a given school.

@@

Despite the use of different educational quality and staff turnover metrics, addition of a relatively extensive set of controls, the estimation strategy proposed here remains observational. In the absence of exogenous shock in either turnover or coalition shares, a credible identification strategy for each set of estimations remains a challenge. To partially address these concerns, as an additional robustness check I replicate these estimations with weights estimated through a covariate balancing propensity score outlined in \citet{imai_covariate_2014} which is generalized for continuous treatments. A covariate balance test for either estimation is presented in the appendix.

@@

\begin{figure}[h]
    \centering
  \includegraphics[width=0.6\linewidth]{plots/plot_coalition_coef_mun.pdf}
  \caption{{\bf Political alignment and staff turnover.} An increase in the share of legislative seats held by the mayoral coalition decrease the amount of turnover for teachers and school principals, including hires or dismissals. Model 1 presents results on the turnover index. Models 2 and 3 are logistic regressions at the individual, bureaucrat level. Models 4 and 5 are aggregated at the municipal level. where the outcome of interest is the proportion of staff either hired or dismissed at a given year. Models 1, 3, and 5 include year and state fixed effects.}
  \label{fig:coalition_turnover}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{plots/plot_interactions_logit}
    \caption{Predicted values for bureaucratic turnover. The graph plots the proportion of educational staff either hired or fired for any given year against the share of seats controlled by the executive. The circles' size is proportional to the number of observations contained at each value of the legislative share of seats.}
    \label{fig:turnover}
\end{figure}

Finally, I show that the executive-legislative bargain extends to second-term mayors. This finding suggests that patronage is not motivated solely by reelection concerns. Rather, it achieves an important policy goal for incumbent mayors who seek to implement their preferred policy.

\begin{figure}
    \centering
    \includegraphics{plots/plot_coalition_coef_mun_second.pdf}
    \caption{\bf{hello}}
\end{figure}

Moving on to characterizing the downstream effects of patronage, I find that teacher and school principal turnover have significant, negative effects on average test scores. These results are robust to alternative specifications of turnover, as well as the use of different quality metrics (SAEB, SPAECE) in our estimations. Table \ref{fig:hlm_mods} presents the results of the estimation on SAEB test scores. We present two specifications for turnover. Turnover index measures the amount of turnover in teachers at the school level. Work experience for teachers and school principals serve as an alternative measure of turnover. Personnel who has only recently entered the school recently amass a shorter amount of work experience (less than two years).

\begin{figure}[h]
  \centering
  \footnotesize
  \includegraphics{plots/model_turnover_learning.pdf}
  \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning. Models 1 and 2 present results for teacher turnover index constructed at the school level. Models 3 and 4 estimate the effect of new teachers and school principals entering the school (less than two years). All models include year and state fixed effects.}
  \label{fig:hlm_mods}
\end{figure}

\blue{Include some reference coefs in order for the audience to assess the magnitude of the effects}

% As a final note, this set of estimations are based on observational data and therefore suffer from well-known concerns of endogeneity. It is in these circumstances that validating causal mechanisms through fieldwork and careful data analysis can increase the validity of a causal claim. In-depth interviews with local actors involved in managing education, as well as the use of different measurements for both educational staff turnover and student learning, provides strong evidence that staff turnover has a negative impact on student learning. Bureaucratic turnover stems from political considerations.

@@

\section{Conclusion}

@@

% # Appendix

% ## Covariate balance

% ### Education quality vs. staff turnover
